>

### 2017-01-05 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Update paper draft

---
title: >
    Working with daily climate model output data in R and the \pkg{futureheatwaves} package
author:
  - name: G. Brooke Anderson
    affiliation: Colorado State University
    address:
    - Department of Environmental & Radiological Health Sciences
    - 1681 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  brooke.anderson@colostate.edu
  - name: Colin Eason
    affiliation: Colorado State University
    address:
    - Department of Computer Science
    - 1873 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  aimesce@gmail.com
  - name: Elizabeth A. Barnes
    affiliation: Colorado State University
    address:
    - Department of Atmospheric Science
    - 1371 Campus Delivery
    - Fort Collins, CO 80523
    email:  eabarnes@atmos.colostate.edu
abstract: >
  Research on climate change impacts can require extensive processing, especially when using ensemble techniques to incorporate output from different climate models and different simulations of each model. This processing can be particularly extensive when identifying and characterizing multi-day extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps. Further, climate model output is in a format and follows standards that may be unfamiliar to most R users. Here, we provide an overview of working with daily climate model output data in R. We then present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output. 
preamble: >
  % Supposedly allows you to add stuff to LaTeX preamble. Not for me.
output: rticles::rjournal_article
---

@@

A key source of data for ensemble techniques is the Coupled Model Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}). This project brought together dozens of major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models. This created an ensemble of state-of-the-art climate model projections that allow researchers to study projections and their uncertainties. Most of these modeling groups additionally performed more than one simulation for each scenario and model (i.e. multiple ensemble members), perturbing the initial conditions by a very tiny amount to quantify uncertainties due to internal climate variability.

@@

For climate impact studies, a main source for climate model output is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5). Over 20 climate modeling groups created one or more climate models which, for this project, were run using standardized scenarios \citep{taylor2012overview}. The resulting output is uniform across modeling groups and has a consistent structure, which allows comparison of simulations from different models \citep{IPCCch9}. CMIP5 climate model output is archived at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple levels in the ocean or atmosphere (e.g., ocean temperature). Here, we will focus on data with a daily time step for variables reported at a single level (e.g., near-surface air temperature).

Each modeling group ran simulations under several experiments, with experiments varying in terms of radiative forcing by using different scenarios of time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.) \citep{taylor2012overview, IPCCch9}. Experiments include historical experiments (run using radiative forcing consistent with observed and reconstructed data for 1850--2005), pre-Industrial control experiments, and experiments of future scenarios of radiative forcing over the 21st century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}. Some modeling groups created ensembles of output for a specific model and experiment, in which they ran the experiment multiple times with the model with very small changes to the initial conditions.

The CMIP5 climate model output data are distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at https://pcmdi.llnl.gov/search/cmip5/. Users must register before downloading data, and some data are restricted to non-commercial use. There is a separate file for each combination of climate model, experiment, modeling realm (e.g., atmosphere, ocean), variable, time step, and ensemble member \citep{taylor2012overview, taylor2010cmip5}. For finer time scales, the output is further split across multiple files for specific year ranges (e.g., 5 years of output for each file) \citep{taylor2010cmip5}. Each file's name includes the output variable, climate model, experiment, and ensemble member for the simulation \citep{taylor2010cmip5}.

CMIP5 files can be searched and downloaded through a point-and-click web interface. They can also be downloaded in bulk to computers with Unix or Mac operating systems using the \code{wget} file downloading utility. Appropriate \code{wget} scripts can be created either through the World Climate Research Programme CMIP5 data portal or through Earth System Grid Federation's Search RESTful API. Tips on efficiently searching and downloading the data, including through use of wget scripts and the search API, are available as user tutorials through the website of the University of Colorado Boulder's Earth System CoG (e.g., https://www.earthsystemcog.org/projects/cog/doc/wget for a tutorial on downloading files using wget).

CMIP5 files are saved in Network Common Data Format (netCDF), a binary file format that allows storage of data representing a regular array. For climate model output at a single level (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two coordinates of location (e.g., latitude and longitude). Figure \ref{fig:netcdfexample} provides a sketch of the structure of netCDF files for single-level climate model output. Global climate models generate output at regularly-spaced time steps, typically at regularly-spaced grid points around the world. The latitude and longitude spacing of grid points vary by climate model, but are typically 1--2 degrees for atmospheric variables in CMIP5 models \citep{IPCCch9}. Each data point in the netCDF array gives the modeled value of the variable (e.g., surface temperature) for a single time point and location. For CMIP5 climate model output, the location units are in degrees east and degrees north for longitude and latitude, respectively. For daily output files, the time unit is in days since a specified origin date-time (e.g., days since 1850-01-01 00:00:00) \citep{taylor2010cmip5}. All CMIP5 output files are required to include certain metadata  \citep{taylor2010cmip5}, including the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}. The metadata also must include units for all of the coordinate variables (e.g., longitude, latitude, time). The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory.

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{netcdf_figure}
\end{center}
\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single level, like near-surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and climate grid location. These data are typically indexed in the netCDF file by longitude, latitude, and time, in that order. For example, if the near-surface air temperature is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{figure}

@@

CMIP5 data are available as netCDF files, for which base R import functions do not exist. There are, however, a few R packages that can be used to work with the netCDF file format used for CMIP5 files. Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN. More recent packages, including \pkg{ncdf4} \citep{ncdf4} and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3. While climate model output data for CMIP5 are required to conform with the earlier version (version 3) \citep{taylor2010cmip5}, it is safer to write code using functions that can be used with either version, in case future phases of CMIP do not require files to conform with netCDF version 3.

You can do a number of things with netCDF files in R using these functions. For example, \pkg{ncdf4}'s \code{nc\_open} function can be used to open a connection to a netCDF file, and the object returned by the function includes the file's attribute data. Once a file connection is open, variables can be read in using the \code{ncvar\_get} function. For example, the variables defining the dimensions of the sketched netCDF file in Figure \ref{fig:netcdfexample} could be read into R with \code{ncvar\_get} with the \code{varid} parameter set to "lat", "lon", or "time". The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}. In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}. If you only need data from a subset of the full file, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file (for example, with the \code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}). Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).

@@

Two R packages provide help with non-Gregorian calendars: \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers}. The \code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in the CMIP5 netCDF file's attributes to convert the "time" variable in the file to an object of the \code{PCICt} class. This class is defined in the \pkg{PCICt} package and provides Date-like functionality for 360- and 365-day calendars \citep{PCICt}. However, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the \code{PCICt} object correctly matches the expected date ranges for a file when processing CMIP5 data in R.

@@

In addition to these general packages for working with netCDF files, there are several R packages specifically for working with climate model output data, including \pkg{RCMIP5} \citep{RCMIP5} and \pkg{wux} \citep{wux}. However, these packages are more useful for working with data output at time steps of a month or higher and have limited utility with the daily climate model output data required for studies of multi-day extreme events.

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special \code{cmip5data} class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename. For this package, most functions only work with monthly or less frequent data \citep{RCMIP5}. While the \code{loadCMIP5} function does successfully load daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data. The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a data frame with information parsed from the file name. As a note, the \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 monthly-aggregated output directly from within R with the \code{CMIP5fromESGF} function. The package then uses the \code{models2wux} function to read in climate model output netCDF files and convert them to "WUX" data frames, which can be used by other functions in the package. While this function can input climate model output with daily time steps (the "what.timesteps" element of the \code{modelinput} list input must be set to "daily"), the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX data frame. Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.

The functions and packages described in this section can be used with CMIP5 netCDF files to do things in R like map near-surface air temperatures from a single climate model on a specific day (Figure \ref{fig:worldmap}) or pull a time series of daily near-surface air temperature simulations at a specific climate model grid point (Figure \ref{fig:timeseries}). The "starting_from_netcdf" vignette that comes with the "futureheatwaves" package provides all code required to create these figures, as well as more details and code examples on working with CMIP5 netCDF files in R.

@@

Once the user creates these data frames of location-specific heat waves, the \code{apply\_all\_models} function can be used to apply custom functions across all the extreme event data frames generated by \code{gen\_hw\_set}. This functionality allows users to create summaries of extreme events across all climate models and ensemble members (Figure \ref{fig:overview}, right). The function can be used to generate summary statistics (e.g., determine average heat wave length or total frost days) or to apply more complex functions (e.g., apply epidemiologic effect estimates across the heat waves to generate health impact estimates).

When using this package, CMIP5 climate model output data require some pre-processing. Data will need to be saved in a specific format, with files stored in a specific directory structure. Full details of the required file and directory structure are provided in the package's "futureheatwaves" vignette, while tips and an R script for conducting this processing starting from CMIP5 netCDF files are given in the "starting_from_netcdf" vignette.

This package can be used for study locations worldwide. The "starting_from_netcdf"" package vignette provides an example of using this package to identify and explore future heat waves in several Chinese cities.

@@

```{r eval = FALSE}
projection_dir_location <- system.file("extdata/cmip5",
                                       package = "futureheatwaves")
city_file_location <- system.file("extdata/cities.csv",
                                  package = "futureheatwaves")

@@

Once the user has created a directory of characterized event files for each ensemble member ("Extreme event data sets", Figure \ref{fig:overview}), he or she can explore the results using the \code{apply\_all\_models} function. This function allows the user to apply custom R functions across all extreme event data frames created by the \code{gen\_hw\_sets} call. The user can apply any R function that follows certain standards in accepting input and returning output. Full details on these standards are given in the \pkg{futureheatwaves} package vignette.

As an example, if the user wanted to calculate the average temperature of the heat waves identified for each ensemble member, he or she could write a simple function:

@@

This output is structured as ``tidy" data \citep{wickham2014tidy}, allowing it to be used easily with the graphing package \CRANpkg{ggplot2} \citep{ggplot2}. This functionality can be used to create a number of other summaries of the identified extreme events. For example, it could be used to determine average heat wave length or estimate how much earlier in the year events are expected to start across an ensemble of climate model simulations. The functionality can also be used for more complex analysis of extreme event files. For example, it can be used to apply epidemiological models of heat wave to estimate excess heat-related mortality under different future scenarios; an example of this application is provided in the "futureheatwaves" vignette that comes with the package.

@@

Therefore, this package allows the user to extensively customize the definition used to identify extreme events. Users can write a custom R function with either a different heat wave definition (see \citet{smith2013heat} and \citet{kent2014heat} for listings of some of the definitions used in scientific studies) or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day spells). For heat wave identification, researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study. For studies of other extreme events, a heat wave definition likely will not be applicable and so a customized definition is necessary.

Three components of the extreme event definition can be easily customized in the \code{gen\_hw\_set} function call, without creating a new R function to use to identify heat waves. First, many extreme event definitions are based on conditions that are rare in the study location \citep{IPCCch1}, but may vary in how rare conditions must be for a period to qualify as an extreme event. For example, some of the different definitions used to identify heat waves vary only in the percentile temperature used for a threshold (e.g., one definition is $\ge2$ days at or above the 98\textsuperscript{th} percentile temperature at a location while another is $\ge2$ days at or above the 99\textsuperscript{th} percentile temperature; \citet{kent2014heat, smith2013heat}). Therefore, the \pkg{futureheatwaves} package allows users to change the percentile of the variable of interest required for an extreme event using the \code{probThreshold} option in \code{gen\_hw\_set}. Other heat wave definitions vary only in the number of consecutive days that must be over the threshold for a period to quality as an extreme event (e.g., one definition is $\ge2$ days at or above the 98\textsuperscript{th} percentile temperature at a location while another is $\ge4$ days at or above the 98\textsuperscript{th} percentile temperature; \citet{anderson2009weather}). Therefore, the package allows the user to change the number of days used in the heat wave definition using the \code{numDays} argument in the \code{gen\_hw\_set} function. Combined, these two customization choices allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of the 16 heat wave definitions used in \citet{kent2014heat} could be fit using different combinations of these two options for specifying threshold percentile and number of days. Third, some extreme events like cold waves and frost day spells are defined as a certain number of days below, rather than above, a threshold. While the default is to identify events by searching for days above a threshold, this behavior can be changed with the \code{above\_threshold = FALSE} argument in the \code{gen\_hw\_set function}.

Beyond these simpler options, the customization of the event definition is even more extensive as one has the option of writing and using a custom R function to identify extreme events. This functionality allows the user to use definitions that either require a number of days above an absolute threshold (e.g., maximum temperature of $\ge$ 95\textsuperscript{o}F for $\ge1$ day \citet{kent2014heat, tan2007heat}; minimum temperature $\le$ 0\textsuperscript{o}C for $\ge1$ day for frost day spells) or that require a combination of thresholds to be met (e.g., maximum daily temperature above a lower threshold every day of the heat wave and above a higher threshold for a certain number of days within the heat wave; \citet{kent2014heat, peng2011toward}). To use a customized event definition, the user must write and load an R function that implements the definition. This custom function is passed to the \code{gen\_hw\_set} function using the \code{IDheatwavesFunction} argument. To work correctly, this custom function must allow only specific inputs and generate only specific outputs; details about the required structure are provided in the \pkg{futureheatwaves} package vignette. To increase processing speed when identifying extreme events, we coded parts of the default event definition function in C++ and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}. Users should consider a similar strategy for custom heat wave definitions, especially if processing a large number of climate projection files.

## Exploring adaptation scenarios

>>>>

---
title: >
    Working with Daily Climate Model Output Data in R and the \pkg{futureheatwaves} Package
author:
  - name: G. Brooke Anderson
    affiliation: Colorado State University
    address:
    - Department of Environmental & Radiological Health Sciences
    - 1681 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  brooke.anderson@colostate.edu
  - name: Colin Eason
    affiliation: Colorado State University
    address:
    - Department of Computer Science
    - 1873 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  aimesce@gmail.com
  - name: Elizabeth A. Barnes
    affiliation: Colorado State University
    address:
    - Department of Atmospheric Science
    - 1371 Campus Delivery
    - Fort Collins, CO 80523
    email:  eabarnes@atmos.colostate.edu
abstract: >
  Research on climate change impacts can require extensive processing of climate model output, especially when using ensemble techniques to incorporate output from multiple climate models and multiple simulations of each model. This processing can be particularly extensive when identifying and characterizing multi-day extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps. Further, climate model output is in a format and follows standards that may be unfamiliar to most R users. Here, we provide an overview of working with daily climate model output data in R. We then present the \CRANpkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output. This package can input a directory of climate model output files, identify all extreme events using customizable event definitions, and summarize the output using user-specified functions.
preamble: >
  % Supposedly allows you to add stuff to LaTeX preamble. Not for me.
output: rticles::rjournal_article
---

@@

A key source of data for ensemble techniques is the Coupled Model Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}). This project brought together major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models. This created an ensemble of state-of-the-art climate model projections that allow researchers to study projections and their uncertainties. Most of these modeling groups additionally performed more than one simulation for each scenario and model (i.e. multiple ensemble members), perturbing the initial conditions by a very tiny amount to quantify uncertainties due to internal climate variability.

@@

For climate impact studies, a main source of data is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5). Over 20 climate modeling groups created one or more climate models which, for this project, were run using standardized scenarios \citep{taylor2012overview}. The resulting output is uniform across modeling groups and has a consistent structure, which allows comparison of simulations from different models \citep{IPCCch9}. CMIP5 climate model output is archived at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple levels in the ocean or atmosphere (e.g., ocean temperature is reported at different ocean depths). Here, we will focus on data with a daily time step for variables reported at a single level (e.g., near-surface air temperature).

Each modeling group ran simulations for CMIP5 under several experiments, with experiments varying in terms of radiative forcing scenarios through the use of different scenarios of time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.) \citep{taylor2012overview, IPCCch9}. Experiments include historical experiments (run using radiative forcing consistent with observed and reconstructed data for 1850--2005), pre-Industrial control experiments, and experiments of future scenarios of radiative forcing over the 21st century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}. Some modeling groups created ensembles of output for a specific model and experiment, in which they ran the experiment multiple times with the model with very small changes to the initial conditions.

The CMIP5 climate model output data are distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at \url{https://pcmdi.llnl.gov/search/cmip5/}. Users must register before downloading data, and some data are restricted to non-commercial use. There is a separate file for each combination of climate model, experiment, modeling realm (e.g., atmosphere, ocean), variable, time step, and ensemble member \citep{taylor2012overview, taylor2010cmip5}. For finer time scales, the output is further split across multiple files for specific year ranges (e.g., 5 years of output for each file) \citep{taylor2010cmip5}. Each file's name includes the output variable, climate model, experiment, and ensemble member for the simulation \citep{taylor2010cmip5}.

CMIP5 files can be searched and downloaded through a point-and-click web interface. They can also be downloaded in bulk to computers with Unix or Mac operating systems using the \code{wget} file downloading utility. Appropriate \code{wget} scripts can be created either through the World Climate Research Programme CMIP5 data portal or through Earth System Grid Federation's Search RESTful API. Tips on efficiently searching and downloading the data, including through use of wget scripts and the search API, are available as user tutorials through the website of the University of Colorado Boulder's Earth System CoG (e.g., \url{https://www.earthsystemcog.org/projects/cog/doc/wget} for a tutorial on downloading files using wget).

CMIP5 files are saved in Network Common Data Format (netCDF), a binary file format that allows storage of data representing a regular array. For climate model output at a single level (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two coordinates of location (e.g., latitude and longitude). Figure \ref{fig:netcdfexample} provides a sketch of the structure of netCDF files for single-level climate model output.

Each data point in the netCDF array gives the modeled value of the variable (e.g., surface temperature) for a single time point and location. Global climate models generate output at regularly-spaced time steps, typically at regularly-spaced grid points around the world. The latitude and longitude spacing of grid points vary by climate model, but are typically 1--2 degrees for atmospheric variables in CMIP5 models \citep{IPCCch9}. For CMIP5 climate model output, the location units are in degrees east and degrees north for longitude and latitude, respectively. For daily output files, the time unit is in days since a specified origin date-time (e.g., days since 1850-01-01 00:00:00) \citep{taylor2010cmip5}.

All CMIP5 output files are required to include certain metadata (or "attributes")  \citep{taylor2010cmip5}, including the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}. The metadata also must include units for all of the dimension variables (e.g., longitude, latitude, time). The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory.

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{netcdf_figure}
\end{center}
\caption{Example of structure of a netCDF climate model output file for a variable reported at a single level, like near-surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and climate grid location. These data are typically indexed in the netCDF file by longitude, latitude, and time, in that order. For example, if the near-surface air temperature in the example netCDF shown here is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{figure}

@@

CMIP5 data are available as netCDF files, for which base R import functions do not exist. There are, however, a few R packages that can be used to work with the netCDF file format used for CMIP5 files. Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN. More recent packages, including \CRANpkg{ncdf4} \citep{CRANncdf4} and \pkg{CRANRNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3. While climate model output data for CMIP5 are required to conform with the earlier version (version 3) \citep{taylor2010cmip5}, it is safer to write code using functions that can be used with either version, in case future phases of CMIP do not require files to conform with netCDF version 3.

You can do a number of things with netCDF files in R using these packages. For example, \pkg{ncdf4}'s \code{nc\_open} function can be used to open a connection to a netCDF file, and the object returned by the function includes the file's attribute data. Once a file connection is open, variables can be read in using the \code{ncvar\_get} function. For example, the variables defining the dimensions of the sketched netCDF file in Figure \ref{fig:netcdfexample} could be read into R with \code{ncvar\_get} with the \code{varid} parameter set to "lat", "lon", or "time".

The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}. In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}. If only a subset of the full file is needed, the dimensional time and location data can be used to identify the location of the needed data in the netCDF array and this information can then be used to read a portion of data into memory (for example, with the \code{nc.get.var.subset.by.axes} function in \CRANpkg{ncdf4.helpers}). Once the user is done reading in data from the file, the connection to the netCDF should be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).

@@

Two R packages provide help with non-Gregorian calendars: \CRANpkg{PCICt} \citep{PCICt} and \CRANpkg{ncdf4.helpers} \citep{ncdf4.helpers}. The \code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in the CMIP5 netCDF file's attributes to convert the "time" variable in the file to an object of the \code{PCICt} class. This class is defined in the \pkg{PCICt} package and provides date-like functionality for 360- and 365-day calendars \citep{PCICt}. However, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the \code{PCICt} object correctly matches the expected date ranges for a file when processing CMIP5 data in R.

@@

In addition to these general packages for working with netCDF files, there are several R packages specifically for working with climate model output data, including \CRANpkg{RCMIP5} \citep{RCMIP5} and \CRANpkg{wux} \citep{wux}. However, these packages are more useful for working with data output at time steps of a month or higher and have limited utility with the daily climate model output data required for studies of multi-day extreme events.

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special \code{cmip5data} class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename. For this package, most functions only work with monthly or less frequent data \citep{RCMIP5}. While the \code{loadCMIP5} function does successfully load daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data. The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a data frame with information parsed from the file name. The \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 output at a monthly time step directly from R with the \code{CMIP5fromESGF} function. The package then uses the \code{models2wux} function to read climate model output netCDF files and convert it to "WUX" data frames, which can be used by other functions in the package. While this function can input climate model output with daily time steps (the "what.timesteps" element of the \code{modelinput} list input must be set to "daily"), the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX data frame. Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data or with output data at a larger time step, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.

The functions and packages described in this section can be used with CMIP5 netCDF files to do things in R like map near-surface air temperatures from a single climate model on a specific day (Figure \ref{fig:worldmap}) or pull a time series of daily near-surface air temperature simulations at a specific climate model grid point (Figure \ref{fig:timeseries}). The \pkg{futureheatwaves}' "Starting from netCDF files" vignette (\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html}) provides all code required to create these figures, as well as more details and code examples on working with CMIP5 netCDF files in R.

@@

Once the user creates these data frames of location-specific extreme events, the \code{apply\_all\_models} function can be used to apply custom functions across all the extreme event data frames. This functionality allows users to create summaries of extreme events across all climate models and ensemble members (Figure \ref{fig:overview}, right). The function can be used to generate summary statistics (e.g., determine average heat wave length or total frost days) or to apply more complex functions (e.g., apply epidemiologic effect estimates across the heat waves to generate health impact estimates).

When using this package, CMIP5 climate model output data require some pre-processing. Data will need to be saved in a specific format, with files stored in a specific directory structure. Full details of the required file and directory structure are provided in the package's main vignette (\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/futureheatwaves.html}), while tips and an R script for conducting this processing starting from CMIP5 netCDF files are given in the "Starting from netCDF files" vignette (\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html}).

This package can be used for study locations worldwide. The "Starting from netCDF files" package vignette provides an example of using this package to identify and explore future heat waves in several Chinese cities.

@@

```{r eval = FALSE}
library(futureheatwaves)
projection_dir_location <- system.file("extdata/cmip5",
                                       package = "futureheatwaves")
city_file_location <- system.file("extdata/cities.csv",
                                  package = "futureheatwaves")

@@

Once the user has created a directory of characterized event files for each ensemble member ("Extreme event data sets", Figure \ref{fig:overview}), he or she can explore the results using the \code{apply\_all\_models} function. This function allows the user to apply custom R functions across all extreme event data frames created by the \code{gen\_hw\_sets} call. The user can apply any R function that follows certain standards in accepting input and returning output. Full details on these standards are given in the main package vignette.

As an example, if the user wanted to calculate the average temperature of the heat waves identified for each ensemble member in the output generated by the code above, he or she could write a simple function:

@@

The same process can be used to create a number of other summaries of the identified extreme events. For example, it could be used to determine average length of extreme events or estimate how much earlier in the year events are expected to start across an ensemble of climate model simulations. The functionality can also be used for more complex analysis of extreme event files. For example, it can be used to apply epidemiological models of heat wave to estimate excess heat-related mortality under different future scenarios; an example of this application is provided in the main \pkg{futureheatwaves} vignette. The output from \code{apply\_all\_models} is structured as ``tidy" data \citep{wickham2014tidy}, allowing it to be used easily with the graphing package \CRANpkg{ggplot2} \citep{ggplot2} and other packages in the tidyverse.

@@

Therefore, this package allows the user to extensively customize the definition used to identify extreme events. Users can write a custom R function with either a different heat wave definition (see \citet{smith2013heat} and \citet{kent2014heat} for listings of some of the definitions used in scientific studies) or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day spells). For heat wave identification, researchers might want to use a different event definition because, for example, it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study. For studies of other extreme events, a heat wave definition likely will not be applicable and so a customized definition is necessary.

Three components of the extreme event definition can be easily customized in the \code{gen\_hw\_set} function call, without creating a new R function to use to identify heat waves. First, many extreme event definitions are based on conditions that are rare in the study location \citep{IPCCch1}, but definitions may vary in how rare conditions must be. For example, some of the different definitions used to identify heat waves vary only in the percentile temperature used for a threshold (e.g., one definition is $\ge2$ days at or above the 98\textsuperscript{th} percentile temperature at a location while another is $\ge2$ days at or above the 99\textsuperscript{th} percentile temperature; \citet{kent2014heat, smith2013heat}). Therefore, the \pkg{futureheatwaves} package allows users to change the percentile of the variable of interest required for an extreme event using the \code{probThreshold} option in \code{gen\_hw\_set}. Other heat wave definitions vary only in the number of consecutive days that must be over the threshold for a period to quality as an extreme event (e.g., one definition is $\ge2$ days at or above the 98\textsuperscript{th} percentile temperature at a location while another is $\ge4$ days at or above the 98\textsuperscript{th} percentile temperature; \citet{anderson2009weather}). Therefore, the package allows the user to change the number of days used in the heat wave definition using the \code{numDays} argument in the \code{gen\_hw\_set} function. Combined, these two customization choices allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of 16 heat wave definitions outlined in \citet{kent2014heat} could be fit using different combinations of these two options for specifying threshold percentile and number of days. Third, some extreme events like cold waves and frost day spells are defined as a certain number of days below, rather than above, a threshold. While the default is to identify events by searching for days above a threshold, this behavior can be changed with the \code{above\_threshold = FALSE} argument in the \code{gen\_hw\_set function}.

Beyond these simpler options, the customization of the event definition is even more extensive as one has the option of writing and using a custom R function to identify extreme events. This functionality allows the user to use definitions that either require a number of days above or below an absolute threshold (e.g., maximum temperature of $\ge$ 95\textsuperscript{o}F for $\ge1$ day \citet{kent2014heat, tan2007heat}; minimum temperature $\le$ 0\textsuperscript{o}C for $\ge1$ day for frost day spells) or that require a combination of thresholds to be met (e.g., maximum daily temperature above a lower threshold every day of the heat wave and above a higher threshold for a certain number of days within the heat wave; \citet{kent2014heat, peng2011toward}). To use a customized event definition, the user must write and load an R function that implements the definition. This custom function is passed to the \code{gen\_hw\_set} function using the \code{IDheatwavesFunction} argument. To work correctly, this custom function must allow only specific inputs and generate only specific outputs; details about the required structure are provided in the main \pkg{futureheatwaves} package vignette. To increase processing speed when identifying extreme events, we coded parts of the default event definition function in C++ and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}. Users should consider a similar strategy for custom heat wave definitions, especially if processing a large number of climate projection files.

## Exploring sensitivity of results to adaptation

Extreme events tend to be identified based on conditions that are rare for a specific location using location-specific relative thresholds. These thresholds are often defined for climate impact studies based on a variable's distribution at that location in present-day or historical data. However, for some extreme events, impacts are associated with how rare the conditions during the event are compared to current norms in the location \citep{anderson2009weather}, which suggests some capacity for adaptation to heat and raises the question of whether extreme events should be defined using a percentile threshold based on present-day variable distributions or based on distributions in the time period being projected. Therefore, it can be interesting to explore trends in extreme events under climate change if extreme events are identified based on variable distributions during the projection period or another future period. The \pkg{futureheatwaves} package allows users to specify the time period to use when determining a location-specific relative threshold for an event definition using the \code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}. This feature allows users to explore how sensitive projections of impacts are to this choice of the time period to use when determining relative variable measures, including thresholds used for percentile-based event definitions.
